\documentclass{article}
\input{../../../../format}
\lhead{Programming Paradigms - Systems Programming}
\usepackage{boxedminipage}
\usepackage{minted}
\definecolor{links}{HTML}{2A1B81}
\usetikzlibrary{shapes.misc}
\usetikzlibrary{shapes.geometric, arrows, positioning}


\begin{document}



\begin{center}
	\underline{\huge Scope and Recursion}
\end{center}




\section{Scope -- where name can be seen}
\begin{minted}{c}
int i;             // i has program scope
                   // and is accessible anywhere

int foo(int j) {   // foo() also has program scope
  int i;           // this i has block scope
                   // and is only accessible between {}
  if (...) {
    int i;         // this i also has block scope
  }
}
static bar() {...} // bar() has file scope
       // and is only accessible by code in this file

float pab(int k);  // k has prototype scope
       // and is only accessible as part of the prototype
\end{minted}



\section{Scope -- where it can be used}

\verb!int !\textcolor{red}{\texttt{i}}\verb!;               !\textcolor{red}{\texttt{i}}\\
\verb!                     !\textcolor{red}{\texttt{i}}\\
\verb!                     !\textcolor{red}{\texttt{i}}\\
\verb!int foo(int j) {     !\textcolor{red}{\texttt{i}}\\
\verb!  int !\textcolor{green}{\texttt{i}}\verb!;               !\textcolor{green}{\texttt{i}}\\
\verb!      !\textcolor{green}{\texttt{i}}\verb!;               !\textcolor{green}{\texttt{i}}\\[\baselineskip]
\verb!  if (...) {           !\textcolor{green}{\texttt{i}}\\
\verb!    int !\textcolor{blue}{\texttt{i}}\verb!;               !\textcolor{blue}{\texttt{i}}\\
\verb!  }                    !\textcolor{green}{\texttt{i}}\\
\verb!}                    !\textcolor{red}{\texttt{i}}\\
\verb!static bar() {...}   !\textcolor{red}{\texttt{i}}\\
\verb!                     !\textcolor{red}{\texttt{i}}\\
\verb!float pab(int k);    !\textcolor{red}{\texttt{i}}

\begin{itemize}
\item Which \verb!i! is visible?
\end{itemize}



\section{Lifetime -- variable birth and death}
\begin{itemize}
\item Three types of lifetime:
\begin{itemize}
\item Static -- life of the program
\item Automatic -- till the end of the current block
\item Dynamic -- we control (\verb!malloc()!/\verb!free()!)
\end{itemize}

\begin{minted}{c}
int* d;
int foo(int j) {
  static int t;                  // static
  int p;                         // automatic
  d = malloc(400*sizeof(int));   // dynamic
}
int bar(int k) {
  free(d);
}
\end{minted}
\end{itemize}



\section{Storage classes}
\begin{itemize}
\item Each variable in C has one of the following four storage types (these are also keywords):

\item \verb!extern! (not the same as \verb!extern! declaration)
\item \verb!static!
\item \verb!auto!
\item \verb!register!
\end{itemize}



\section{extern}
\begin{itemize}
\item When a variable is defined it is allocated storage
\begin{itemize}
\item possibly initialised (\verb!int i = 5;!)
\end{itemize}
\item When a variable is declared it informs the compiler that a variable of a given type exists
\item Top-level variables default to \verb!extern! storage class
\begin{itemize}
\item including definition and declaration
\item but not the \verb!extern! keyword
\end{itemize}
\item Use \verb!extern! keyword to declare but not define a variable
\begin{itemize}
\item i.e. it will be defined elsewhere but accessible here
\end{itemize}
\item Lifetime and scope of whole program
\item Cf. abstract classes and interfaces in Java
\end{itemize}



\section{\texttt{extern} keyword}
\begin{itemize}
\item Use a variable from a different file
\end{itemize}
\begin{tabular}{lll}
\texttt{func.c} &  & \texttt{main.c}\\
\begin{boxedminipage}{5.3cm}
\begin{minted}{c}
int cost;

int compute_cost(int q) {
  return q * cost;
}




\end{minted}
\end{boxedminipage}
&&
\begin{boxedminipage}{5cm}
\begin{minted}{c}
#include <stdio.h>
extern int cost;
int compute_cost(int q);
int main() {
  cost = 5;
  printf("cost = %d\n",
     compute_cost(3));
  return 0;
}
\end{minted}
\end{boxedminipage}
\end{tabular}

\begin{itemize}
\item To run:
\begin{minted}{c}
gcc -c func.c
gcc main.c func.o
./a.out
\end{minted}
\end{itemize}



\section{\texttt{static}}
\begin{itemize}
\item \verb!static! and \verb!extern! are mutually exclusive as keywords
\item \verb!static! variables have the same lifetime as the program - so on calling a function it will keep the same value as was defined before
\item \verb!static! global variables (i.e. those outside function declarations)
 have \emph{file scope}
\item \verb!static! local variables (i.e. those inside function declarations) have \emph{function scope}
\item Calling a variable \verb!static! is confusing because it means different things in different languages
\begin{itemize}
\item and also within C
\end{itemize}
\end{itemize}



\section{\texttt{auto}}
\begin{itemize}
\item Automatic variables have the same lifetime as the function in which they are defined
\item They have function scope
\item Automatic variables are stored in the \emph{stack frame}
\item Local variables are automatic by default, so the \verb!auto! keyword is never explicitly used in practice.
\item (\verb!auto! was part of C from the early days to make it easier to convert code from B, where it was necessary when defining local variables. \emph{N.B.} \verb!auto! has a very different meaning in C++!)
\end{itemize}



\section{\texttt{register}}
\begin{itemize}
\item Suggests that a variable should (if possible) be stored in a register rather than in main memory
\item Cannot use the address of (\verb!&!) operator on register variables
\item Storing in a register is much faster to access
\item Not all register variables are necessarily stored in registers 
\begin{itemize}
\item may be too many
\end{itemize}
\item Not all variables stored in registers are declared as such 
\begin{itemize}
\item code optimisation
\end{itemize}
\item Modern compilers are very good at working out which variables are best made into register variables and will do this in the background automatically, so using \verb!register! is quite rare
\end{itemize}



\section{Local variables}
Properties of local variables
\begin{itemize}
\item Automatic storage duration:
\begin{itemize}
\item Storage is automatically allocated when the function is called and de-allocated when it terminates
\end{itemize}

\item Block scope:
\begin{itemize}
\item A local variable is visible from its point of declaration to the end of the enclosing function body
\item These are stored in the function context on the call stack
\end{itemize}
\item In performance terms they do add a small overhead to each function call
\end{itemize}



\section{Example Stack}
\begin{itemize}
\item The stack is an area of memory used for temporary storage
\item Often (but not always) used for
\begin{itemize}
\item Return addresses
\item Local variables
\item Parameters
\item Return values
\end{itemize}
\end{itemize}
\begin{tabular}{cc}
\begin{minipage}{7cm}
\begin{minted}{c}
int function(int p1,
     int p2, int p3) {
  int A, B, C;
  ... 
}
\end{minted}
\end{minipage}
&
\begin{tabular}{|c|}
\hline
Variable \verb!A!\\
\hline
Variable \verb!B!\\
\hline
Variable \verb!C!\\
\hline
Return address\\
\hline
Parameter \verb!p1!\\
\hline
Parameter \verb!p2!\\
\hline
Parameter \verb!p3!\\
\hline
Some other value\\
\hline
\end{tabular}
\end{tabular}




\section{Call stack example}
\usetikzlibrary{decorations.pathreplacing,angles,quotes}
\begin{tikzpicture}

\node[align=left] at (-4,1) {Stack pointer};
\draw[->] (-2.7,1) -- (-1.4,1);
\node[align=left] at (-4,0) {Frame pointer};
\draw[->] (-2.7,0) -- (-1.4,0);

\node[align=left] at (-4,-2.5) {\texttt{DrawSquare} has\\ called \texttt{DrawLine}};
%Stack frame for\\ \texttt{DrawLine} function
%Stack frame for\\ \texttt{DrawSquare} function

\draw[fill=lime] (-1.3, 0) rectangle (1.3, 1) {};
\node[align=center] at (0,0.5) {Locals of\\\texttt{DrawLine}};
\draw[fill=lime] (-1.3, 0) rectangle (1.3, -0.5) {};
\node[align=center] at (0,-0.25) {Return address};
\draw[fill=lime] (-1.3, -1.5) rectangle (1.3, -0.5) {};
\node[align=center] at (0,-1) {Parameters of\\\texttt{DrawLine}};

\draw[decoration={brace,mirror,raise=3pt},decorate] (1.3,-1.5) -- (1.3,1);
\node[align=left] at (3.3,-0.25) {Stack frame for\\ \texttt{DrawLine} function};

\begin{scope}[shift={(0,-2.5)}]
\draw[fill=cyan] (-1.3, 0) rectangle (1.3, 1) {};
\node[align=center] at (0,0.5) {Locals of\\\texttt{DrawSquare}};
\draw[fill=cyan] (-1.3, 0) rectangle (1.3, -0.5) {};
\node[align=center] at (0,-0.25) {Return address};
\draw[fill=cyan] (-1.3, -1.5) rectangle (1.3, -0.5) {};
\node[align=center] at (0,-1) {Parameters of\\\texttt{DrawSquare}};

\draw[decoration={brace,mirror,raise=3pt},decorate] (1.3,-1.5) -- (1.3,1);
\node[align=left] at (3.3,-0.25) {Stack frame for\\ \texttt{DrawSquare} function};
\end{scope}
\end{tikzpicture}



\section{Code block scope}
\begin{itemize}
\item Block scope refers to any code block not just functions
\begin{minted}{c}
if (a > b) {
  int tmp = a;
  // tmp is local to this code block

  a = b;
  b = tmp;
}
\end{minted}
\item \verb!tmp! is automatic and local
\end{itemize}



\section{Static and global variables}
\begin{itemize}
\item \verb!static! variables exist for the duration of the program
\item Variables declared outside a function are visible to all code in the same program and are \verb!static! by default
\begin{minted}{c}
// scope inside a single source file
int a = 10;       // global & static
static int c = 1; // file & static

foo(){
  int tmp = 3;    // local automatic
  static int count = 0; // local static
  a = a + tmp;
  count++;
}
\end{minted}

\item Same \verb!count! variable each time you call \verb!foo()!
\end{itemize}



\section{Function parameters}
\begin{itemize}
\item Parameters have the same properties as local variables
\begin{itemize}
\item i.e. automatic storage duration and block scope
\item Each formal parameter is initialized automatically when a function is called (by being assigned the actual value of the corresponding argument)
\end{itemize}
\end{itemize}



\section{Summary of scope in a single file}
\begin{itemize}
\item \verb!file1.c!:
\end{itemize}
\begin{minted}{c}
int gv;               // gv - global scope (static)

static int fv;        // fv - file scope (static)

void f( int pv ){     // pv - block scope of f()
                      //    (automatic)

   int lv = 0;        // lv - block scope (automatic)

   static int sv = 0; // sv - blck scope (static)
}
\end{minted}



\section{Pros and cons of global variables}
\begin{itemize}
\item Global variables are convenient when many functions must share a variable or when a few functions share a large number of variables
\item In most cases, it's better for functions to communicate through parameters rather than shared variables:
\begin{itemize}
\item If we change a global variable during program maintenance (by altering its type, say), we'll need to check every function in the same file to see how the change affects it
\item If a global variable is assigned an incorrect value, it may be difficult to identify the guilty function
\item Functions that rely on global variables are hard to reuse in other programs
\end{itemize}
\end{itemize}



\section{Iterative functions}
\begin{minted}{c}
int loop_power( int a, int n ) {
  int result = 1;
  while (n > 0 ) {
    result = result * a;
    n--;
  }
  return result;
}
\end{minted}

\begin{itemize}
\item Calculate \verb!a! raised to the power \verb!n!
\item \verb!1 * a * a *!  $\cdots$ \verb!n! times
\end{itemize}



\section{Recursive functions}
\begin{itemize}
\item Recursion is an alternative to using a loop

\item C allows this by allowing functions to call themselves

\item Like any loop this needs:
\begin{itemize}
\item initial conditions
\item conditional test (a termination test)
\item a variable change, e.g. a decrement
\end{itemize}

\item Relies on a new function scope being created every time a function calls itself
\end{itemize}



\section{Recursive power function}
\begin{minted}{c}
int recursive_power(int a, int n){
  if ( n == 0 )  // termination test
    return 1;    // base case
  else {         // recursive case 
    return ( a * recursive_power( a, n - 1 ) );
  }
}
\end{minted}

\begin{itemize}
\item The loop variable here is \verb!n!, which decrements to zero as repeated recursive calls are made
\item N.B. Repeated need to initialise the function context is a cost
\end{itemize}



\section{Recursive power function -- short form}
\begin{minted}{c}
int recursive_power2( int a, int n ) {
  return (n == 0) ? 1 : (a * recursive_power2(a,n-1));
}
\end{minted}

\begin{itemize}
\item The same function written with the conditional operator:
\begin{minted}{c}
value = expr1 ?  expr2 : expr3;
\end{minted}

\item is the same as
\begin{minted}{c}
if (expr1)
  value = expr2;
else
  value = expr3;
\end{minted}
\end{itemize}



\section{Factorials}
\begin{itemize}
\item Look at \verb!factorial.c! (a little simpler than \verb!power.c!)

\item \verb!gcc -g -Wa,-ahl=factorial.s factorial.c!
\item This will interleave assembly language with source code statements

\item If you know the way function calls are made, you can mix C programs with assembly language programs
\end{itemize}



\section{Recursive Fibonacci function}
\begin{minted}{c}
int fib(int n) {
  if(n >= 2)      // recursion test  
    return fib(n-1) + fib(n-2); // recursive case
  return n;       // base case for n = 0 and n =1
}
\end{minted}

\begin{itemize}
\item This returns the \verb!n!th element in the Fibonacci series:

\item \verb!0,1,1,2,3,5,8,13,21,! etc.

\item Base case: \verb!fib(0)! is \verb!0! and \verb!fib(1)! is \verb!1! 
\item Production rule: \verb!fib(n) = fib(n-1) + fib(n-2)!
\end{itemize}



\section{Recursion summary}
\begin{itemize}
\item Write loops using a function that calls itself
\item This must have both a:
\begin{itemize}
\item base case
\item recursive case
\end{itemize}
\item To terminate, the base case must happen

\item Relies on the run time system to:
\begin{itemize}
\item create the function's scope
\item keep track of the local variables for each call
\item This is a performance overhead compared to iterative loops
\end{itemize}
\end{itemize}

\end{document}
