\documentclass{article}[18pt]
\input{../../../../format}
\lhead{MODULE}
\usepackage{hyperref}
\usepackage{minted}

\providecommand{\tightlist}{%
	\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\begin{document}
\begin{center}
\underline{\huge Coursework Questions}
\end{center}


\hypertarget{assignment-questions}{%
	\section{Assignment questions}\label{assignment-questions}}

\hypertarget{assignment-questions-1}{%
	\subsection{Assignment questions}\label{assignment-questions-1}}

\begin{itemize}
	\tightlist
	\item
	What does a reusable component look like?
	\item
	What is a \texttt{draw} method with optional p5.Renderer as parameter?
	\item
	When are private fields appropriate?
	\item
	How to use ESLint?
	\item
	Can I convert a processing sketch?
	\item
	Others?
\end{itemize}

\hypertarget{what-does-a-reusable-component-look-like}{%
	\subsection{What does a reusable component look
		like?}\label{what-does-a-reusable-component-look-like}}

\begin{itemize}
	\tightlist
	\item
	Variables in sketch become properties of the object
	\item
	Setup code in sketch becomes the constructor
	\item
	Draw method in sketch becomes draw method of the object
	\item
	Other methods for changing parameters and handling events
\end{itemize}

\hypertarget{minimal-usage-in-sketch-index.js}{%
	\subsection{Minimal usage in sketch
		(index.js)}\label{minimal-usage-in-sketch-index.js}}

Include p5 library your sketch class definition

\begin{minted}{js}
var c;

function setup(){
	c = new Component();
}

function draw(){
	c.draw();
}
\end{minted}

\hypertarget{what-is-a-draw-method-with-optional-p5.renderer-as-parameter}{%
	\subsection{\texorpdfstring{What is a \texttt{draw} method with optional
			p5.Renderer as
			parameter?}{What is a draw method with optional p5.Renderer as parameter?}}\label{what-is-a-draw-method-with-optional-p5.renderer-as-parameter}}

\begin{itemize}
	\tightlist
	\item
	p5.Renderer objects are returned by \texttt{createCanvas} and
	\texttt{createGraphics}
	\item
	Can be used e.g.~as texture for another object
	\item
	All calls to graphics e.g.~\texttt{rect}, \texttt{ellipse},
	\texttt{fill} can be applied
\end{itemize}

\begin{minted}{js}
function draw(g){
	if(g){
		g.ellipse(this.x, this.y, this.d, this.d);
	}
	else{
		ellipse(this.x, this.y, this.d, this.d);
	}

}
\end{minted}

To avoid duplication you could

\begin{itemize}
	\tightlist
	\item
	(Optionally) create the canvas in the constructor and save the return
	value
	\item
	OR create delegation functions inside your draw function e.g.
\end{itemize}
\newpage
\begin{minted}{js}
rect(x,y,w,h){
	if(g){
		g.rect(x,y,w,h);
	}
	else{
		rect(x,y,w,h);
	}
}
\end{minted}

\hypertarget{when-are-private-fields-appropriate}{%
	\subsection{When are private fields
		appropriate?}\label{when-are-private-fields-appropriate}}

\begin{itemize}
	\tightlist
	\item
	They are a a post-ES6 experimental feature
	\item
	On reflection, probably not appropriate to use
\end{itemize}

\hypertarget{how-to-use-eslint}{%
	\subsection{How to use ESLint?}\label{how-to-use-eslint}}

\begin{itemize}
	\tightlist
	\item
	Want to use \href{https://eslint.org/docs/rules/}{eslint:recommended}
	rules
	\item
	\href{https://eslint.org/docs/user-guide/getting-started}{Set up on
		your machine}
	\item
	When running \texttt{eslint\ -\/-init} choose
	\texttt{Answer\ questions\ ...}
\end{itemize}

\begin{verbatim}
DA-ECS-01:ball dcs0spb$ eslint --init
? How would you like to configure ESLint? Answer questions about your style
? Which version of ECMAScript do you use? ES2015
? Are you using ES6 modules? No
? Where will your code run? Browser
? Do you use CommonJS? No
\end{verbatim}

\begin{verbatim}
? Do you use JSX? No
? What style of indentation do you use? Spaces
? What quotes do you use for strings? Single
? What line endings do you use? Unix
? Do you require semicolons? Yes
? What format do you want your config file to be in? JavaScript
\end{verbatim}

You might want to use Windows line endings

\hypertarget{eslint-output}{%
	\subsection{ESLint output}\label{eslint-output}}

\begin{verbatim}
DA-ECS-01:ball dcs0spb$ eslint ball.js

/Users/dcs0spb/Documents/Prog18potatoes/ball/ball.js
1:7   error  'Ball' is defined but never used                    no-unused-vars
3:1   error  Expected indentation of 8 spaces but found 1 tab    indent
4:1   error  Expected indentation of 8 spaces but found 1 tab    indent
5:1   error  Expected indentation of 8 spaces but found 1 tab    indent
5:24  error  Unnecessary semicolon                               no-extra-semi
\end{verbatim}

\hypertarget{can-i-convert-a-processing-sketch}{%
	\subsection{Can I convert a processing
		sketch?}\label{can-i-convert-a-processing-sketch}}

Yes

\begin{itemize}
	\tightlist
	\item
	Replace variable types with \texttt{var} (or \texttt{let} or
	\texttt{const})
	\item
	Replace method return types e.g.~\texttt{void} with \texttt{function}
	\item
	Find equivalent objects e.g.~\texttt{new\ PVector} becomes
	\texttt{createVector}
\end{itemize}

\hypertarget{others}{%
	\subsection{Others?}\label{others}}



\end{document}