\documentclass{article}[18pt]
\input{../../../../format}
\lhead{New Venture Creation}


\begin{document}
\begin{center}
\underline{\huge Effectual Entrepreneurship}
\end{center}
2 readings at the end, important reading about writing a good business plan. - red and amber readings are the best
\section{Casual vs Effectual Thinking}
\begin{itemize}
	\item Casual Thinking — Aligning all resources towards a goal.
	\begin{itemize}
		\item This is the "outside in" way of thinking.
		\item Line up the resources towards the group of customers you are aiming for.
		\item An example of this is easy jet - no frills
	\end{itemize}
	\item Effectual Thinking
	\begin{itemize}
		\item Split means out into many imagined ends
		\item "inside out" way of thinking
		\item An example of this is Honda - Known competency in engines.
	\end{itemize}
\end{itemize}
Principles of effectuation:
\begin{itemize}
	\item The bird in the hand principle
	\begin{itemize}
		\item Starting with a modest idea allows you to explore putting more into your idea.
	\end{itemize}

	\item The affordable loss principle
	\begin{itemize}
		\item Not betting what you don't have, only put in what you can afford to lose
		\item Related to bootstrapping
	\end{itemize}

	\item The crazy quilt principle
	\begin{itemize}
		\item 
	\end{itemize}

	\item The lemonade principle
	\begin{itemize}
		\item 
	\end{itemize}

	\item The pilot in the plane principle
	\begin{itemize}
		\item 
	\end{itemize}
\end{itemize}
\section{Lean Startup}
Lean startup - Looking at the minimum amount an entrepreneur can do to benefit the customer. This allows the entrepreneur to start engaging with the customers, who start to give feedback on what is being done
\begin{itemize}
	\item Begin with the idea
	\item Measure how well you are doing through customer testing
	\item Learn and implement what you need to improve
\end{itemize}
\section{Bootstrapping}
There are many ways to fund the business without borrowing lots of money.
\begin{itemize}
	\item Get money from clients before committing any resources.
	\item Look for a deposit big enough to get the resources needed.
	\item Negotiate late payment with supplier
	\item Many more examples
\end{itemize}

\end{document}