\documentclass{article}[18pt]
\usepackage{../../../../format}
\lhead{Theory of Computation - Algorithms and Complexity}


\begin{document}
\begin{center}
\underline{\huge Greedy Algorithms}
\end{center}
\section{The knapsack problem}
We consider two variants:
\begin{enumerate}
	\item The \textbf{0-1 knapsack problem}: A thief robs a store and finds n items. Item $i$ is worth $v_i$ euros and weighs $w_i$ kilos.The thief can carry at most W kilos in knapsack.\\
	\textit{Question}: Which items should they take in order to maximise their profit?
	\item The \textbf{fractional knapsack problem}: Same set-up, but this time the thief can carry fractions of items
\end{enumerate}
We assume that the knapsack is not big enough to carry all the items
$$W< w_1+...+w_n$$
\subsection{Greedy Strategy}
\textbf{Fractional} knapsack problem is solvable by following \textbf{greedy} strategy
\begin{enumerate}
	\item Compute the value per kilo $r_i=v_i/w_i$ for each item
	\item Sort items by value per kilo in nonincreasing order
	\item Take as much as possible of item with greatest value per kilo, then take as much as possible of item with the next greatest value per kilo etc. until total weight equals knapsack capacity W
\end{enumerate}

\begin{theo}[nonincreasing]{thm:nonincreasing}
Nonincreasing allows for items of the same weight, unlike decreasing 
\end{theo}
\subsection{Running time of greedy strategy}
Step 1 (computing $r_i$ for each $i$) can be done in $\Theta(n)$\\
\\
Step 2 (sorting the $r_i$ array) can be done in $\Theta(n\log n)$\\
\\
Step 3 (selecting the items) can be done in $\Theta(n)$\\
\\
Thus, the worst-case running time is $\Theta(n\log n)$
\subsection{Greedy strategy for 0-1 knapsack}
The same algorithm does NOT work for the 0-1 knapsack problem\\
\\
For instance W=6 and 
\[
\begin{array}{lll}{v_{1}=3,} & {w_{1}=4,} & {r_{1}=3 / 4} \\ {v_{2}=2,} & {w_{2}=3,} & {r_{2}=2 / 3} \\ {v_{3}=2,} & {w_{3}=3,} & {r_{3}=2 / 3}\end{array}
\]
\textbf{Optimal solution}: steal objects 2 and 3 for a total of 4 euros\\
\textbf{Greedy strategy}: steal object 1 for a total of 3 euros
\subsection{Correctness for fractional knapsack}
n items are sorted by $r_i=v_i/w_i$ in nonincreasing order. So
$$r_i \geqslant r_j \quad \text{for any} \quad i<j$$
Two items $p$ and $q$ with $r_p=r_q$ are considered to be equivalent. Hence, we may assume that
$$r_i > r_j \quad \text{for any} \quad i<j$$
The total available amount of item $i$ is $w_i$. Let k be such that
\[
w_{1}+\ldots+w_{k} \leq W \quad \text { and } \quad w_{1}+\ldots+w_{k+1}>W
\]
(Recall that we assume that the knapsack is not big enough to carry all items, hence such a $k$ exists)\\
\\
Let $x=(x_1,...,x_n)$ denote a solution: $x_i$ is the amount the thief takes of item $i$ for $i=1,...,n$\\
The total value stolen is then $\sum_{i=1}^{n}r_ix_i$\\
\\
The greedy solution $x$ is given by
\[
\begin{aligned} x_{i} &=w_{i} \text { for } 1 \leq i \leq k \\ x_{k+1} &=W-\left(w_{1}+\ldots+w_{k}\right) \\ x_{i} &=0 \quad \text { for } k+2 \leq i \leq n \end{aligned}
\]
Let $x*$ be an optimal solution. First of all, we must have $x^*_1+...+x^*_n=W$ (We might as well fill up the sack!)\\
\\
We will show that $x^*=x$\\
\\
Suppose $x^*\neq x$. Then $x^*_i<x_i$ for some $i\leqslant k+1$: let $j := min\{i: x^*_i<x_i\}$. Also, $x^*_I>x_I$ for some $I>j$\\
\\
Let us add more of item $j$ to the optimal choice: Let $y$ be another solution, defined by
\[
\left\{\begin{array}{l}{y_{j}=\min \left\{w_{j}, x_{j}^{*}+x_{I}^{*}-x_{I}\right\}} \\ {y_{I}=x_{I}^{*}-y_{j}+x_{j}^{*}} \\ {y_{i}=x_{i}^{*} \quad \text { otherwise }}\end{array}\right.
\]
Exercise:
\begin{enumerate}
	\item Check that $y$ is a valid solution, i.e. $0\leqslant y_i \leqslant w_i$ for all $1\leqslant i \leqslant n$ and that $y_1+...+ y_n=W$
	\item Check that $y$ is a better solution than $x^*$, i.e. $\sum_{i=1}^{n}r_iy_i>\sum_{i=1}^{n}r+ix^*_i$. Contradiction
\end{enumerate}
\section{Greedy Algorithms}
The knapsack problem is an example of an optimization problem. In an optimization problem there can be many solutions. Each solution has a value. An optimal solution is a solution with optimal (minimum or maximum) value. We want to find an optimal solution\\
\\
Algorithms for optimization problems go through sequence of steps. At each step there is a set of choices. 
\begin{theo}[Greedy Algorithm]{thm:greedy}
A greedy algorithm makes in each step the choice that looks the best at the moment. This choice is called the greedy choice or locally optimal choice.
\end{theo}
\subsection{Performance of greedy algorithms}
This strategy does not have to lead to optimal solutions. Example: 0-1 knapsack problem.\\
\\
For several problems solution obtained by greedy algorithm is optimal. Examples:
\begin{itemize}
	\item Fractional knapsack problem
	\item Minimum spanning tree (Prim, Kruskal)
	\item Shortest paths from a single source (Dijkstra)
	\item Data compression (Huffman coding)
\end{itemize}


\end{document}